---
title: "Exercise Prediction - PML Project"
author: "Philip Mateescu"
date: "January 29, 2016"
output: html_document
---
```{r echo=FALSE,message=FALSE,warning=FALSE}
library(dplyr)
library(caret)
library(pgmm)
library(rpart)
library(gbm)
library(lubridate)
library(forecast)
library(e1071)
library(randomForest)
library(gridExtra)
```

# Using Machine Learning to Predict Exercise Correctness 

Using fitness device (*Jawbone Up*, *Nike FuelBand*, *Fitbit*) 
[data collected](http://groupware.les.inf.puc-rio.br/har) by a group 
of enthusiasts, we will attempt to construct a Machine Learning
model that will predict "how (well)" an activity has been performed
by the weared.

The dataset has been obtained from the [Human Activity Recognition](http://groupware.les.inf.puc-rio.br/har) project.

## Dataset Cleanup and Exploratory Analysis

The training data has been obtained from
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
and the test data has been downloaded from
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

Both files have been places in a `data` subdirectory.

```{r}
training <- read.csv('data/pml-training.csv')
training$classe <- as.factor(training$classe)
testing <- read.csv('data/pml-testing.csv')
```

From the study's [documentation](http://groupware.les.inf.puc-rio.br/public/papers/2013.Velloso.QAR-WLE.pdf), we learned that the authors have added eight feature
that are calculate values of other variables: mean, variance,
standard deviation, max, min, amplitude, kurtosis and skewness.  
We will eliminate these.

```{r}
calculatedColumns <- grepl('^(mean|var|stddev|max|min|skew|kurt|ampli|avg)', colnames(testing))
```

Upon initial inspection using `str(testing)` and `summary(testing)`
we notice that there is no data (either N/A or empty) for some of the 160 variables in the dataset.
These would *not* make for good predictor variables.

```{r}
isMissing <- function(x) { any(is.na(x) || x == "")}
missingValues <- sapply(testing, isMissing)
```

We will combine the two sets to create our predictor set and exclude them
alongside some of the columns that don't hold measurement data (e.g. 
*user_name* or timestamps)

```{r}
excluded <- c('X', 'user_name', 'new_window', 'num_window', 'cvtd_timestamp',
              'problem_id', 'raw_timestamp_part_1', 'raw_timestamp_part_2')
n <- names(testing)
n <- n[!missingValues | !calculatedColumns]
predictors <- n[!(n %in% excluded)]
predictors
```

These predictors include both "features on the Euler angles (roll, pitch
and yaw), as well as the raw accelerometer, gyroscope and
magnetometer readings" ^[1]^.

Let's create three sets of predictors: one that contains only raw data,
one that contains the angles, and one that contains both and see
which one yields better results.

```{r}
pred.angles <- predictors[grepl('roll|pitch|yaw', predictors)]
pred.raw <- predictors[!grepl('roll|pitch|yaw', predictors)]
training <- training[,c('classe', predictors)]
```


As we already have a testing set, let's split the training set
into a training and a validation set. Since the testing set 
is very small, we'll create a larger validation set.

```{r}
inTrain <- createDataPartition(training$classe, p = 0.6)
data.train <- training[inTrain[[1]], ]
data.validation <- training[-inTrain[[1]], ]
```
Distribution of classe observations in the original distribution vs the 
`data.train` and `data.validation`:
```{r echo=F}
g.sec <- function(g) {
    g + geom_bar() + 
        guides(fill=F)  + 
        theme(axis.text.y=element_blank(), axis.ticks.y=element_blank(), axis.title.y=element_blank())
}
g1 <- ggplot(training, aes(x=classe, fill=classe)) + geom_bar() + theme(legend.position='none') + ggtitle('Training (original)')
g2 <- g.sec(ggplot(data.train, aes(x=classe, fill=classe))) + ggtitle("Training")
g3 <- g.sec(ggplot(data.validation, aes(x=classe, fill=classe))) + ggtitle("Validation")
grid.arrange(g1, g2, g3, ncol=3)

```






# References and Citations


1. Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.